Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-15 main.Rmd
add abstract

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm_latex.tex
    toc: no
    includes:
      in_header: preamble.tex
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: "TBD."
geometry: margin=1in
biblio-style: apsr
bibliography: ["r-references.bib"]
graphics: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{siunitx}
  - \usepackage{float}
---

@@

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by hyperparamter scans. All models were implemented in Python 3.6.9 using the deep learning API Keras 2.2.4 with Tensorflow 1.12.0 backend, using GPU NVIDIA GK110GL Tesla K20c and CPU AMD Opteron(TM) Processor 6272 with memory 128 GB.

@@

Hyperparameter scans were performed using random search, a hyperparameter optimization method which 
implements a randomized search over hyperparameters, where each hyperparameter setting is randomly sampled from a distribution over possible hyperparameter values [@Bergstra2012]. In particularly, we used RandomSearchCV(), the implementation of random search method in `scikit-learn 0.19.2` [@scikit-learn] together with KerasClassifier() which is implementation of the scikit-learn classifier API for keras. Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). All model training was performed as described in section Model Training.

@@

We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding. To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 3 * 4 * 50 * 5` models in total. ^[For each architecture, 50 models are selected randomly. Therefore there are 3 taxon * 4 architecture * 50 models * 5 cross validation]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam). For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction. ^[\@ref(tab:possiblecombination) is table of total number of all hyperparameter settings in the whole hyperparameter space for each architecture] From this we had twelve candidate models to develop further.

@@

Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised and although only a small fraction ^[This fraction depends for every model, the parameter space is shown in table \@ref(tab:possiblecombination)] of the possible hyperparameter and parameter space was explored we still compared `r 3 * 4 * 50 * 5` models. At a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding. So even for relatively small data sets, a large amount of specialised GPU compute power is required.

>>>>

---
title: "Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm_latex.tex
    toc: no
    includes:
      in_header: preamble.tex
fontsize: 12pt
fontfamily: sans
papersize: a4paper
author:
  - name: Ruth Kristianingsih, Dan MacLean
  - affiliation: The Sainsbury Laboratory, University of East Anglia, Norwich Research Park, Norwich, UK, NR4 7JJ, dan.maclean@tsl.ac.uk
abstract: "Plant pathogens cause billions of dollars of crop loss every year and are a major threat to global food security. Effector proteins are the tools such pathogens use to infect the cell, predicting effectors de novo from sequence is difficult because of the heterogeneity of the sequences. We hypothesised that deep learning classifiers based on Convolutional Neural Networks would be able to identify effectors and deliver new insights. We built a training set of manually curated effector sequences from PHI-Base and used these to train a range of model architectures for classifying bacteria, fungal and oomycete sequences. The best performing classifiers had accuracies from 93 % to 84 %. The models were tested against popular effector detection software on our own test data and data provided with those models. We observed better performance from our models. Specifically our models showed greater accuracy and lower tendencies to call false positives on a secreted protein negative test set and a greater generalisability. We used GRAD-CAM activation map analysis to identify the sequences that activated our CNN-LSTM models and found short but distinct N-terminal regions in each taxon that was indicative of effector sequences. No motifs could be observed in these regions but an analysis of amino acid types indicated differing patterns of enrichment and depletion that varied between taxa. We have produced an R package that will allow others to make easy effector predictions using our models."
geometry: margin=1in
biblio-style: apsr
bibliography: ["r-references.bib"]
graphics: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{siunitx}
  - \usepackage{float}
---

@@

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by hyperparamter scans. All models were implemented in Python 3.6.9 [@python] using the deep learning API Keras 2.2.4 [@chollet2015] with the Tensorflow 1.12.0 [@tensorflow2015] backend, using  NVIDIA GK110GL Tesla K20c GPUs and AMD Opteron(TM) Processor 6272 CPUs with 128 GB RAM.

@@

Hyperparameter scans were performed using random search, a hyperparameter optimization method where each hyperparameter setting is randomly sampled from a distribution of possible hyperparameter values [@Bergstra2012]. We used RandomSearchCV(), the implementation of random search in `scikit-learn 0.19.2` [@scikitlearn] together with KerasClassifier() which is an implementation of the scikit-learn classifier API for keras. Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). All model training was performed as described in section Model Training.

@@

We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding. To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 3 * 4 * 50 * 5` models in total. Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam). For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction. From this we had twelve candidate models to develop further.

@@

Training the models proved to be computationally expensive, the architectures used have a large number of parameters and hyperparameters to be optimised (see Table \@ref(tab:possiblecombination) for a breakdown) and although only a small fraction  of the possible hyperparameter and parameter space was explored we compared `r 3 * 4 * 50 * 5` models, at a run time of around 144 minutes for CNN-LSTM, 57.2 minutes for CNN-GRU, and 45 minutes for both GRU Embedding and LSTM Embedding. For these relatively small data sets, a significant amount of specialised GPU compute power was required.
